\textsc{Solving the heat-equation using neural networks}

### 2019-12-20 assignment_three/article/main.tex
Cleaned up. Done with project 3.

\setlength{\parskip}{1em}
\setlength{\parindent}{0em}
\usepackage[]{eulervm} 
\usepackage[]{palatino} 
\usepackage[]{commath} 
\usepackage[]{microtype} 
\usepackage[]{geometry} 
\usepackage[]{minted} 
\usemintedstyle{algol}

@@

\title{\textsc{Solving the heat-equation using neural networks}}
\author{Ivar Stangeby}

@@

@@

	In order to evaluate our numerical approximations, whether they are
	computed using finite-difference or they are trained neural networks,
	we have several metrics we can consider. Calling our numerical
	approximation to \( u \) for \( \tilde{u} \), we define our errors as
	\begin{equation}
		e_p = \| u - \tilde{u} \|_p,
	\end{equation}
	where \( \| \cdot \|_p \) denotes the \(p\)-norm for vectors.

@@

\clearpage
\printbibliography
\end{document}

>>>>

\setlength{\parskip}{1em}
\setlength{\parindent}{0em}
\usepackage[]{eulervm} 
\usepackage[]{palatino} 
\usepackage[]{graphicx} 
\usepackage[]{commath} 
\usepackage[]{booktabs} 
\usepackage[]{microtype} 
\usepackage[]{geometry} 
\usepackage[]{minted} 
\usemintedstyle{algol}

@@

\newcommand*{\github}[1]{\def\@github{\color{gray!70}#1}}
\newcommand{\githublogo}{\raisebox{-1pt}{\includegraphics[height=9pt]{imgs/github}}}
\title{\textsc{Solving the heat-equation using neural networks}}
\author{Ivar Stangeby}

@@

	\begin{abstract}
		In this project we compare a standard finite difference method
		to neural networks in the solution of the heat equation. We
		show that the neural network severely underperforms when
		compared to the finite differences, and that the run-time is
		significantly slower. This gives us no immediate reason to
		chose a network based approach over classical methods. All
		relevant code can be found on GitHub:

		\centering{\href{https://github.com/qTipTip/FYS-STK4155}{\githublogo \, \color{gray!50}
github.com/qTipTip }}

	\end{abstract}

@@

	In order to evaluate our numerical approximations, whether they are
	computed using finite-difference or they are trained neural networks,
	we have several metrics we can consider. Calling our numerical
	approximation to \( u \) for \( \tilde{u} \), we define our errors as
	\begin{equation}
		e_p = \| u - \tilde{u} \|_p,
	\end{equation}
	where \( \| \cdot \|_p \) denotes the matrix \(p\)-norm. We consider in
	particular the \( \infty\)-norm, coinciding with the maximum absolute
	error, and the \( 2 \)-norm, which is the standard euclidean
	(Frobenius) norm.

@@

	We start by running the forward in time--central in space scheme. The
	errors are reported in \cref{tab:ftsc_errors}. The 2-norm steadily
	decreases, which means that on average the approximation becomes better
	with decreased grid-size. Also, the \( \infty\)-norm decreases steadily
	towards zero. With this in mind, the approximations are fairly good at
	even low spatial resolutions. Due to the stability criterion, running
	this method for even smaller grid-sizes in space quickly becomes
	infeasible.

	\begin{table}[htpb]
		\centering
		\caption{Solving the heat equation for varying spatial
			resolution, with the temporal resolution chosen to
			satisfy the stability criterion. We report both the \(
			e_1 \) and \( e_2 \) errors.} 
		\label{tab:ftsc_errors}
		\begin{tabular}{lrr}
			\toprule
			\textsc{FTCS} & \( e_\infty \) & \( e_2 \) \\
			\midrule
			\( \Delta x = 1 / 10 \) & \num{0.00357} & \num{0.04674} \\
			\( \Delta x = 1 / 20 \) & \num{0.00098} & \num{0.03688} \\
			\( \Delta x = 1 / 30 \) & \num{0.00044} & \num{0.03113} \\
			\( \Delta x = 1 / 40 \) & \num{0.00025}& \num{0.02743} \\
			\( \Delta x = 1 / 100 \) & \num{0.00004} &\num{0.01787} \\
			\bottomrule
		\end{tabular}
	\end{table}

	We now run the neural network. During testing we realized that the
	neural network also needs a balance between the number of data-points
	in the temporal domain and the number of data-points in the spatial
	domain. We therefore run the network with both spatial and temporal
	resolution increasing at the same rate, following that of
	\cref{tab:ftsc_errors}. We evaluate the same errors. The network is run
	for \( 10000 \) epochs (which in hindsight might be on the fewer end of
	things), and the results are shown in \cref{tab:nnpde_errors}. Note in
	particular how the error not neccesarily decreases with higher
	resolution. This might indicate that a different architecture is to be
	prefered. Maybe a shallower network with larger width. This will also
	reduce the run-time of the network, as the current architecture is
	quite sluggish.

	Taking the reported errors at face-value, the classical finite
	difference method is to prefer, despite its stability issues.

	\begin{table}[htpb]
		\centering
		\caption{Solving the heat equation for identical spatial and
			teporal resolution.  In addition to the errors not
			neccesarily decreasing on increased resolution, the
			network complexity quickly increases, which incurs
			significant run-time increase. This might indicate that
			a different architecture is to be prefered.
			} 
		\label{tab:nnpde_errors}
		\begin{tabular}{lrr}
			\toprule
			\textsc{FTCS} & \( e_\infty \) & \( e_2 \) \\
			\midrule
			\(\Delta t = \Delta x = 1 / 10 \) & \num{0.03684} & \num{0.14905} \\
			\(\Delta t =  \Delta x = 1 / 20 \) & \num{0.14728} & \num{1.24897} \\
			\(\Delta t =  \Delta x = 1 / 30 \) & \num{0.03979} & \num{0.39177} \\
			\(\Delta t =  \Delta x = 1 / 40 \) & \num{0.04963}& \num{0.55036} \\
			\(\Delta t =  \Delta x = 1 / 100 \) & \num{0.05103} &\num{1.35165} \\
			\bottomrule
		\end{tabular}
	\end{table}

	\chapter{Conclusion}

	In this project we have discussed the solution of the heat-equation,
	both using finite-difference methods, and a neural network based
	approach. Based on the results, the \emph{forward in time---centered in
	space} difference scheme outperformed the neural net both in accuracy
	and in runtime, despite its stability issues.

	The benefits of using a neural network for this problem is hard to see
	at face value. One benefit is however that the neural network does not
	need an underlying domain discretization. This does not constitute any
	advantage in such a simple case as our 1D diffusion equation, but in
	higher dimensions, this can prove to be a great advantage.

	For instance in the finite element method, the numerical properties of
	the approximation greatly hinges on the quality of the underlying mesh.

	The poor results of the neural network sparked an interest in trying
	out different architectures, however I spent too much time trying to
	make this work using \texttt{PyTorch}, and realized too late that I had
	to migrate to \texttt{Tensorflow}, where my experience is lacking.

	In conclusion, for simple tasks like this, it seems like the neural
	network works well, comparatively, for very coarse domain
	discretizations, however falls behind finer resolutions.

	Outside of the scope of this project would be the solution of the 2D
	wave equation using a neural network over a non-trivial domain, which
	might be able to highlight some of the benefits of not requiring domain
	discretization.
\clearpage

\appendix

\chapter{Finding eigenvalues}

In this appendix we briefly discuss the idea behind how you can estimate
eigenvalues and eigevectors of a real symmetric matrix using the neural network
framework discussed in this project. Unfortunately, I was unable to make the
network converge, and due to time-constraints this was relegated to an
appendix.

Following \cite{yiNeuralNetworksBased2004} we can also use our neural network
to find the eigenvalues of a real symmetric matrix \( A \). In particular, let
\( Q \)  be a \( 6 \times 6 \) random matrix and define
\begin{equation}
	A = \frac{Q^T + Q}{2}, 
\end{equation}
which then is a symmetric real matrix.

In \cite{yiNeuralNetworksBased2004} they show that any non-zero equilibrium
point of the system
\begin{equation}
	\pd{u(t)}{t} = -u(t) + f(u(t)), 
\end{equation}
where \( f(u) \) is defined as
\begin{equation}
	f(u) = \left[u^Tu A + \left(1 - u^TAu\right)I\right]u
\end{equation}
is an eigenvalue of the matrix \( A \).
Here \( u \) is the network output.

To recast this into our framework, we wish to minimize the function 
\begin{equation}
	\pd{u(t)}{t} + u(t) - f(u(t))
\end{equation}
with respect to our network parameters. The network is shown to converge to
\emph{an} eigenvalue of the matrix \( A \). However, you can make the network
converge to the largest eigenvalue by choosing an initial condition for the
network non-orthogonal to the eigenspace of the largest eigenvalue. This
eigenspace is not known a-priori, so a random initialization is made with the
motivation that the probability is rather large that the random vector is not
orthogonal to the eigenspace.
\printbibliography
\end{document}
